\documentclass{article}
\usepackage[utf8]{inputenc}

\pagestyle{plain}
\setlength{\parindent}{0pt}
\setlength{\parskip}{1ex plus 0.5ex minus 0.2ex}
% command for speaker turns...
% copy this into your preamble if you copy parts of the transcript into another document:
\newcommand{\turn}[2]{
\item[#1:] #2
}
\begin{document}
\begin{center}\textbf{salt}\end{center}

\begin{description}

\turn{ADAL-Examiner}{I'm at Byron kindergarten on ((what date is it)) the 30th of March 2021?\\
My name is X[REDACTED]. and I'm here with X[REDACTED], doing the oral language assessment. [CENSOR]\\
Okay, so now Ada it's your turn to tell the story.\\
You can look at the pictures when you're telling the story.\\
So let's start at the beginning.\\
What was the story about?}

\turn{ADAL-Child}{(Um the kids) the kid/s, \{ in-situ comment\}  they quickly put their gumboot/s on.}

\turn{ADAL-Examiner}{save/ed mud/y bush/s bus/s put/ing girl/z[EP:boy/z] want/3s go/3s shop/ing run/ing drop/ed help/ed aunty/z stop/ed leave/ing come/ing it/z lift/*ed.\\
Anything else?}

\turn{ADAL-Child}{And please go for a walk? :01 [EU]\\
You need to put your gumboot/s on.\\
It/'s too dark\^{} }

\turn{ADAL-Examiner}{What happened in this one?}

\turn{ADAL-Child}{And then it/'s heaps\_and\_heaps|heaps dark.}

\turn{ADAL-Examiner}{What happened next?}

\turn{ADAL-Child}{\%yip\_yip Schnitzel\_von\_Krumm s* :02 falled|fall[EW] out *of the baby/s/z nest.}

\turn{ADAL-Examiner}{:03 What happened next?}

\turn{ADAL-Child}{They put (them) it back in the nest (um).\\
Bye bye little bird.}

\turn{ADAL-Examiner}{Anything else that happened?}

\turn{ADAL-Child}{x.}


\end{description}

\end{document}
