\documentclass{article}
\usepackage[utf8]{inputenc}

\pagestyle{plain}
\setlength{\parindent}{0pt}
\setlength{\parskip}{1ex plus 0.5ex minus 0.2ex}
% command for speaker turns...
% copy this into your preamble if you copy parts of the transcript into another document:
\newcommand{\turn}[2]{
\item[#1:] #2
}
\begin{document}
\begin{center}\textbf{salt}\end{center}

\begin{description}

\turn{ADAL-Examiner}{I'm at Byron kindergarten on what date is it the 30th of March 2021?\\
My name is \_. and I'm here with \_, doing the oral language assessment.\\
Okay, so now Ada it's your turn to tell the story.\\
You can look at the pictures when you're telling the story.\\
So let's start at the beginning.\\
What was the story about?}

\turn{ADAL-Child}{Um the kids the kids, they quickly put their gumboots on.}

\turn{ADAL-Examiner}{saved muddy bushes buses putting girl's wants goes shopping running dropped helped aunty's stopped leaving coming its lift.\\
Anything else?}

\turn{ADAL-Child}{And please go for a walk?\\
You need to put your gumboots on.\\
It's too dark\^{} }

\turn{ADAL-Examiner}{What happened in this one?}

\turn{ADAL-Child}{And then it's heaps and heaps dark.}

\turn{ADAL-Examiner}{What happened next?}

\turn{ADAL-Child}{Schnitzel von Krumm s\~{}  falled out the babies' nest.}

\turn{ADAL-Examiner}{What happened next?}

\turn{ADAL-Child}{They put them it back in the nest um.\\
Bye bye little bird.}

\turn{ADAL-Examiner}{Anything else that happened?}

\turn{ADAL-Child}{\_.}


\end{description}

\end{document}
